Estimation of the perceived time of presence of sources in urban acoustic environments using deep learning techniques

### 2019-10-02 paper/paperV2.tex
CL/PA proof + Cover

The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution). In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources at a sufficient accuracy. To validate this assertion, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes allows the computation of a physical indicator approximating the perceived time of presence of sources, then used for the training and validation of the proposed estimation procedure. The resulting model predicts the time of presence with excellent performance, which translates to the estimation of pleasantness in urban sound environments at a sufficient degree of precision.

@@

%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/EC European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for noise reduction planning purposes. In this context, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as the "acoustic environment as perceived or experienced and/or understood by people, in context" provides a shift from the noise-only paradigm by considering the possible positive influence of some the sound environment's components.

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}. This requires the identification of its composition in terms of sources of interest from acoustic measurements at a reasonable cost. Most monitoring applications mainly rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013}, which offers limited information regarding the content of the sound environment. Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio on much finer time scales, in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}. While using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators, it may not be desirable in long-term monitoring applications where analysis is performed offline. Indeed, it requires large transmission bandwidth and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved. Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information. This representation is easier to store in the long term and reduces privacy concerns~\cite{gontier2017}. In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes. Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomies, it demonstrates that the information content of third octave spectra is sufficient to identify sound sources.

The characterization of urban soundscapes through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. Three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}. Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010} or the time of presence~\cite{ricciardi2014}. The perceived time of presence has the advantage of considering all simultaneously active sources, and its identification from acoustic monitoring measurements would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which has demonstrated state-of-the-art performance in many tasks studied in the machine listening community~\cite{mesaros2017}. In the context of urban sound environments, deep learning has successfully been applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017} and sound scene classification~\cite{valenti2016, salamon2017-2}. Several architectures may be used to this aim, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}. Learning-based approaches mainly process spectral representations of the audio, such as Mel spectrograms or Mel frequency cepstrum coefficients~\cite{barchiesi2015}. Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with labels consisting of perceived source activity. It is possible to manually annotate each recording in the corpus. However, this process would be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources. As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes. By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs. Thanks to this data generation procedure, we are able to validate the proposed approach.

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes to a sufficient degree of accuracy through a comparative study with state of the art approaches.
\end{enumerate}

The procedure for simulating sound scenes corpora is presented in Section~\ref{sec:data} with the outcomes of the listening test conducted to justify the relevance of the sound scenes corpora in the context of a study on perceived quality. A large corpus is constructed and used to train and evaluate a deep learning architecture that performs multi-label sound source recognition at relevant time scales, as presented in Section~\ref{sec:pred}. The application of the resulting time of presence to the prediction of pleasantness in urban environments is then presented in Section~\ref{sec:app}.%The, in comparison to existing models from perceptual and physical indicators.

@@

Considering machine learning techniques to detect the presence of sources requires the availability of annotated data. Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed. This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality~\cite{lavandier2006}. The corresponding database of isolated is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. \textit{simScene} generates original scenarios using the following information:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions,
\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
\end{itemize}

Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. To do so, a corpus of 74 recordings obtained in~\cite{aumond2017} is taken as reference. The recordings include 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. The statistics on background and event properties needed for \textit{simScene} generation are extracted conditionally on the ambiance from annotations of the corpus presented in~\cite{gloaguen2017}, which include:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and their respective sound level considered constant,
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Furthermore, since voice events in the isolated source database consist mostly of read English recordings, voice event mean EBR are reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes. At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.

@@

The generation procedure is the same for both datasets. Though, it is important in deep learning applications to ensure a complete independance of the training and test set. To do so, the isolated samples database is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, while the generation process is the same for both corpora, the scenarios and audio samples are different.

@@

An experiment in the form a listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as obtain subjective annotations for the design and evaluation of a time of presence estimation model.

Given the aim of this test, a separate corpus is constructed. First, 200 simulated scenes of 45s each with diverse new scenarios are generated with the procedure described in Section~\ref{sec:data_corp}, though with a separate database of isolated source occurences. These scenes are equally distributed among the five considered ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected. The distribution of selected extracts is shown in Figure~\ref{fig:tvb_pres}. Playback sound levels are realistically drawn as described in Section~\ref{sec:data_corp} and range from 46.6~dB SPL to 77.1~dB SPL over the 75 sound scenes.

\begin{figure*}[th]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
\end{figure*}

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The annotations obtained in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}. Furthermore, the original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.

The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.\\

During the test, participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. The perceived time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study. The influence of traffic on perception is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test (note that with 23 subjects, the incomplete block design is not perfectly balanced). All gave consent prior to the experiment and reported normal hearing.

@@

The perceptual responses of the 6 recorded and corresponding replicated scenes are first studied. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

@@

Next, the perceptual space generated by the experiment's five general scales (P, L, OL, I and C) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes. These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments. This further demonstrates the effectiveness of the scene generation procedure in terms of diversity of scenarios.

@@

Based on previous studies~\cite{aumond2017, gontier2018, ricciardi2014}, several indicators are identified that correlate well with perceptual parameters. These indicators may be used to automatically annotate sound scenes in the design of a deep learning architecture for sound source recognition.

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level. Typically this includes indicators derived from sound level measurements. For this study the following variables are computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is often associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
\end{itemize}

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:

>>>>

The impact of urban sound on human beings has often been studied from a negative point of view (noise pollution). In the two last decades, the interest of studying its positive impact has been revealed with the soundscape approach (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence, dominance or volume.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources at a sufficient accuracy. To validate this assertion, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. From the contributions of the individual sound sources available for the simulated corpus, a physical indicator approximating the perceived time of presence of sources is computed and used to train and evaluate a multi-label source detection model. This model predicts the presence of simultaneously active sources from fast third octave spectra, allowing the estimation of perceptual attributes such as pleasantness in urban sound environments at a sufficient degree of precision.

@@

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to characterize the sound environment in a way that is closer to the perception of city dwellers~\cite{iso2014}. This requires the identification of its composition in terms of sources of interest from acoustic measurements at a reasonable cost. Most current monitoring applications rely on the measurement of sound levels on time scales from 1s to several hours~\cite{can2008, brocolini2013, nilsson2006}, which offers limited information regarding the content of the sound environment. Conversely, sound environment recognition and event detection applications operate on spectral representations of recorded audio such as Mel spectrograms or Mel frequency cepstrum coefficients, on much finer time scales in tens of milliseconds~\cite{lunden2016, aucouturier2007, cakir2015}. Using such information-rich representations of the signal also allows the computation of commonly used monitoring acoustical indicators. Though, it may not be desirable in long-term monitoring applications where analysis is performed offlinen as it requires large transmission bandwidth and storage capabilities, and raises privacy concerns if intelligible speech can be retrieved~\cite{gontier2017}. Some monitoring applications~\cite{aumond2017, torija2013, nilsson2007} use third octave sound level fast (125ms) measurements to underline relevant spectral information. This representation is easier to store in the long term and reduces privacy concerns~\cite{gontier2017}. In~\cite{aumond2017}, the authors successfully linked acoustical indicators derived from third octave sound levels to the perceived activity of specific sound sources in recordings of polyphonic sound scenes. Though the discriminative properties of the proposed indicators are limited and the underlying methodology cannot be easily extended to larger sound source taxonomies, it demonstrates that the information content of third octave spectra is sufficient to identify sound sources.

The characterization of urban soundscapes through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. A two-dimensional soundscape model of perceived quality emerges where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. In the case of pedestrians, three major source types are found to contribute to these dimensions in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, though the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, brown2011}. Furthermore, traffic noise, human voices and birds calls can respectively be used as a proxy of these source types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017, axelsson2010}, the time of presence~\cite{ricciardi2014}, or the volume. The dominance is built on coupled notions of time and volume, while the time of presence is more easily linked to the physical activity of sources. In a more holistic approach of soundscape quality evaluation, the time of presence should thus be easier to identify from acoustic monitoring measurements and would provide more complete information on the content of polyphonic urban environments~\cite{lavandier2016}.

In this paper, we do so by developing new indicators relying on source recognition models based on deep learning techniques, which has demonstrated state-of-the-art performance in many tasks studied in the machine listening community~\cite{mesaros2017}. In the context of urban sound environments, machine listening has been successfully applied to sound event detection~\cite{foggia2015, mcfee2018, adavanne2017, boes2018}, sound scene classification~\cite{torija2013, barchiesi2015, sun2019, valenti2016, salamon2017-2} and soundscape quality evaluation~\cite{lunden2016, yu2009, boes2018}. A wide range of architectures exist, the most common of which are convolutional and recurrent neural networks~\cite{mesaros2019}. Large amounts of data with task-specific annotations are required for deep learning architectures to learn to extract relevant information from recordings~\cite{salamon2014, mesaros2017, gemmeke2017}. To the best of our knowledge such databases do not currently exist in the literature in the case of source recognition with labels consisting of perceived source activity. It is possible to manually annotate each recording in the corpus. However, this process would be time-consuming, and must be repeated to extend the taxonomy of relevant sound sources. As an alternative, we consider the use of simulated corpora~\cite{lafay2016, salamon2017}, of which the generation procedure provides complete knowledge about the composition of sound scenes. By using a simple indicator computed on this information-rich data to approximate the source-specific perceived time of presence labels, the excerpts composing large corpora may be automatically annotated without the need for additional subjective inputs. Thanks to this data generation procedure, we are able to validate the proposed approach.

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled and perceptual judgements are available\footnote{Corpus available at \url{https://zenodo.org/record/3248734\#.XQjC4v7gqUk}}.
  \item Propose numerical means\footnote{Open source code is available at \url{https://github.com/felixgontier/soundSourcePresenceEstimation}.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches, with a larger corpus of simulated scenes\footnote{Corpus available at \url{https://zenodo.org/record/3248703\#.XQjDVv7gqUk}}.
  \item Demonstrate that the proposed method can be applied to the prediction of a perceptual descriptor of soundscapes (pleasantness) to a sufficient degree of accuracy through a comparative study with state of the art approaches.
\end{enumerate}

Section~\ref{sec:data} presents the generation procedure for a large corpus of simulated sound scenes, its validation through a listening test in the context of a study on perceived soundscape quality, and a study of available acoustical indicators for the automatic annotation of source activity. In Section~\ref{sec:pred} a deep learning architecture that performs multi-label sound source recognition at relevant time scales is trained and evaluated. Section~\ref{sec:app} then presents an application of this model to the prediction of pleasantness in urban environments.

@@

Considering machine learning techniques to detect the presence of sources requires the availability of annotated data. Thus, in order to train deep learning architectures for source recognition, a large corpus is constructed. This corpus is composed of sound scenes simulated using the \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to isolated occurences of these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. Here we choose to restrict the taxonomy of active sources to \textit{traffic}, \textit{human voices} and \textit{birds}, as these sources are found to be the most influent on soundscape quality for pedestrians~\cite{lavandier2006}. The corresponding isolated samples database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. \textit{simScene} generates original scenarios using the following information:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions,
\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
\end{itemize}

Scenarios generated by \textit{simScene} should cover most real-life situations while remaining perceptually plausible. To guarantee the realism of created sound scenes, the information needed for \textit{simScene} generation is based on a corpus of 74 recordings from the GRAFIC project~\cite{aumond2017}. These 74 sound scenes are obtained during 4 soundwalks in 19 locations in the 13th district of Paris, France, and range from 55~s to 4.5~m in duration. In~\cite{gloaguen2017}, the extracts are manually annotated and classified in terms of in terms of ambiance (\textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}). The available annotations include:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and their respective sound level considered constant,
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio in dB.
\end{itemize}

This information is reused to extract statistics on background and event sources activity, which are then used as inputs by \textit{simScene}. Though, adjustments are made on the variance of the considered properties to extend the range of covered scenarios. Furthermore, since voice events in the isolated samples database consist mostly of read English recordings, the voice event mean event-to-background ratios are reduced for all ambiances to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data, with predominant voice activity. Diverse new scenarios are then generated by sampling the background and event properties distributions and used with samples randomly selected from the database of isolated occurrences to simulate sound scenes. At the end of the simulation process the sound level of each sound scene is randomly sampled and conditioned to the ambiance according to typical sound levels in urban context.

@@

The generation procedure is the same for both datasets. Though, it is important in deep learning applications to ensure a complete independance of the training and evaluation sets. To do so, the isolated samples database from which \textit{simScene} assemble extracts to generate a sound scene is split in the same proportions as the two datasets: two-thirds for the development set and the remaining one-third for the evaluation set. As a result, the two corpora differ by the scenarios and isolated samples used in their generation.

@@

An listening test is conducted in order to validate the perceptual correspondence of the proposed corpus with urban sound scene recordings, as well as to obtain subjective annotations for the design and evaluation of the time of presence estimation model. Given the aim of this test, a separate corpus is constructed. This corpus contains 100 sound scenes, including 75 simulated scenes as well as 6 recorded and 19 replicated scenes from the GRAFIC project.

First, 200 simulated scenes of 45s each with diverse new scenarios are generated using the procedure described in Section~\ref{sec:data_corp}, though with a third isolated samples database to ensure independance from the deep learning datasets. These scenes are equally distributed among the five considered ambiances (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}). For each scene the perceived time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected. Figure~\ref{fig:tvb_pres} shows the distribution of the 75 selected extracts. Playback sound levels are realistically drawn for each scene as described in Section~\ref{sec:data_corp}, and range from 46.6~dB SPL to 77.1~dB SPL over the 75 simulated scenes.

\begin{figure*}[th]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with scenarios generated by \textit{simScene}.}\label{fig:tvb_pres}
\end{figure*}

In order to validate the generation procedure of new scenarios, the listening test corpus is extended with replications of recorded sound scenes. Reference recordings are obtained from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations (noted P1-19) with diverse environments. For each of the 19 corresponding recordings, 45 seconds of audio in a single channel are extracted. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The manual annotations available in~\cite{gloaguen2017} in terms of background and event information are then used to replicate the sound scenes using \textit{simScene}. Furthermore, the original 45~s recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are added to the experiment corpus to evaluate changes in perception yielded by the replication process, as they explore diverse real-life situations with respect to ambiance categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\

During the test, participants are asked to evaluate sound scenes in terms of 8 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, tranquil} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. Additionally, to assess the perceived source activity 3 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Time of presence of traffic, voice and bird sources: \textit{Never - Continuously} (\textit{Jamais - Continuellement}) - resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

@@

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test, all reported normal hearing (note that with 23 subjects, the incomplete block design is not perfectly balanced). All participants gave written consent prior to the experiment, and evaluations were further anonymized.

@@

The effect of the \textit{simScene} generation procedure on perception is first investigated. To do so, perceptual responses are compared for the 6 recorded scenes and 6 corresponding replicated scenes, which share common scenarios. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the first five scales, all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Though, overall no consistent difference between the perception of recorded and replicated scenes emerges for the studied points.

@@

Next, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied to validate the use of simulated sound scenes with new scenarios as well as reduced source complexity. It is obtained by performing a principal components analysis on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the principal components space for each scene and represented in Figure~\ref{fig:pspace_sim} as dots for simulated scenes corpus and crosses for recorded and replicated scenes. These projections show that the space covered by scenes based on new scenarios covers that of the studied real-life environments. This further demonstrates the effectiveness of the scene generation procedure in terms of diversity of scenarios.

@@

\begin{figure*}[t!]
    \centering
    \includegraphics[width=1.5\columnwidth]{figures/pred_ex.png}
    \caption{Example of simulated sound scene with ground truth source contributions and resulting presence predictions by the deep learning architecture (greyed out areas)}\label{fig:pred_ex}
\end{figure*}

To train and evaluate a deep learning architecture for sound source recognition using the corpus described in Section~\ref{sec:data_corp}, the activity of sound sources present in each scene must be known. Though, the manual annotation of such information is impractical: it is extremely time-consuming due to the size of the corpus, and the process must be repeated for each studied sound source. Several indicators are identified in the literature that that correlate well with perceptual parameters~\cite{aumond2017, gontier2018, ricciardi2014}. These indicators may be useful to automatically annotate the deep learning corpus without the need for additional human input.
